From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-04-17 paper/paper.Rmd
submitted 2nd revisions

---
title: "From noise to knowledge: how randomness generates novel phenomena and reveals information"
author:
  - name: "Carl Boettiger"
    affiliation: a
    email: "cboettig@berkeley.edu"

address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"

## Use trends style to quickly count refs
#csl: trendsjournals.csl
csl: ecology-letters.csl
bibliography: refs.bib
output: 
  rticles::elsevier_article:
    keep_tex: true
layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
preamble: |
  \newcommand{\ud}{\mathrm{d}}
  \usepackage[nomarkers]{endfloat}
  \journal{Ecology Letters}
  \linenumbers
  \usepackage{setspace}
  \doublespacing

abstract: |
  # Article Type: Review & Synthesis
  # Running title: From noise to knowledge
  # Figures: 4
  # Tables: 0
  # Word count: 7430
  # Word count in abstract: 181
  # Number of References: 111
  # Archive statement:

  All data (simulated data and code) is made available on GitHub (<https://github.com/cboettig/noise-phenomena>), and if accepted, will be archived with provided DOI in an appropriate scientific data repositoriy.

@@

#keywords:
#  - stochasticity
#  - demographic noise
#  - environmental noise
#  - colored noise
#  - quasi-cycles
#  - tipping points

@@

@@

In arguing for this shift, it essential to reconize this is a call for a bigger tent,
not for the rejection of previous paradigms.  What I will characterize as 'noise the nuisance'
reflects a predominately statistical approach, in which noise, almost by definition,
represents all the processes we are not interested in that create additional variation
which might obscure the pattern of interest.  By contrast, an extensive literature has
long explored how noise itself can create patterns and explain processes from population
cycling to coexistence.  These broad categories should be seen as a spectrum and not be
mistaken for either a sharp
dichotomy nor a reference to a stictly empirical-theoretical divide.  Each paradigm
expands upon rather than rejects the previous notion of noise: the recognition that
noise can create novel phenomena does not mean that noise cannot also obscure the
signal of some process of interest.  Likewise, seeking to use noise as a novel source
of information about underlying processes will be informed by both previous paradigms,
as our discussion will illustrate.

Accompanying this discussion, I provide concise and
commented code for simulating each of the models we will discuss as
Appendix A, while more mathematical background and derivations can be found in Appendix B.
Numerical simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.

@@

It is often suggested that stochasticity is inherent or fundamental,
e.g. demographic stochasticity arises because the birth of an individual is probabilistic
[@Melbourne2012]. It is easy to forget that this is merely an assertion about a
model, not an assertion about reality [@Bartlett1960].  Only in quantum mechanics do we find
inherently probabilistic properties: births and deaths of organisms all have
far more mechanistic explanations which we simply summarize in statistical terms:
on average -- that is, summing over a wide range of possible mechanisms we have chosen
not to explicitly model -- an individual death occurs at rate $\lambda$.
This is in fact an extrinsic factor (variables not modeled explicitly),
which makes birth appear probabilistic; (just as we describe the toss of a coin
as probabilistic when in fact it is a mechanistic outcome of Newton's laws).
Demographic stochasticity arises only subsequently, when we change scales from
a description at the individual level to one at the population level.

@@

Before we follow that path, we will examine how this canonical model can emerge from a lower-level description and how it can deviate from that description. This approach will also set the stage to see how our understanding of stochastic models can be extended to better reflect natural processes that shape not only for the so-called deterministic skeleton, but the noise as well.  In so doing, I will highlight two tools fundamental to this analysis, one computational and one analytical: the exact simulation method of @Gillespie1977's stochastic simulation algorithm (SSA), and the van Kampen system size expansion [@vanKampen1976].

@@

The @Gillespie1977 provides an exact algorithm
for simulating demographic stochasticity at an individual level.
The algorithm is merely a simple and direct implementation of the master
equation, progressing in random step sizes determined
by the waiting time until the next event.
Free from both the approximations and mathematical complexity,
the Gillespie algorithm is an interesting example of where we rely on a
numerical implementation to check the accuracy of an analytic approximation,
even in the case of simple models such as we will discuss.  Though the 
algorithm is often maligned as numerically demanding, it can be run much
more effectively even on large models on today's computers than when
it was first developed in the 70s, and remains an underutilized approach
for writing simple and approximation-free^[that is, free from the approximation
made by SDE models as we see in the van Kampen example.  All models are, of course,
only approximations.] stochastic ecological models.

@@

We choose this model because it has the same functional form
as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
but is explicit about how this net growth rate is divided between
birth and death rate.  This leads to real differences in observed stochastic properties
of the model. Both the numerical approach (details in Appendix A) and the 
analytical approach (Appendix B) illustrate that it is straight-forward
to use this same approach in alternate formulations.
Figure \ref{gillespie} shows the results of two exact SSA simulations of this model with
identical parameters except for the total number of available sites,
$N$, illustrating the magnitude of these fluctuations indeed scales in proportion
to $\sqrt{N}$ as postulated by the canonical equation, Eq \eqref{canonical}.

@@

In the interests of a theoretical synthesis, I observe that this result
is *different* than the canonical equation, Eq \eqref{canonical}, or more generally
the diffusion approximation [or Kramer-Moyal expansion @Kramers1940; @Moyal1949; @Gardiner2009]
which is much more typical in the ecological literature [e.g. @Nisbet1982; @Lande2003; 
@Ovaskainen2010]. @Black2012 gives an excellent introduction 
to the van Kampen expansion as a connection between individual-based and population-level
model descriptions and applications, but fails to point out these differences,
mistakenly characterizing the van Kampen expansion as synonymous with the diffusion approximation.
In brief, the diffusion approximation does not lead to a Gaussian distribution
(equivalently, an SDE or PDE with linear noise term, see Appendix B), but rather
a *non-linear* SDE of the form (e.g. @Ovaskainen2010)

@@

The van Kampen expansion can also be useful in illustrating how extrinsic
factors such as environmental noise enter into a stochastic model.
Like the expansion of demographic noise, this derivation leads to 
environmental noise that is approximately Gaussian in distribution,
but unlike the canonical model, leads to explicitly auto-correlated noise 
(see derivation, Appendix B).  Auto-correlated or colored environmental
noise has been a topic of significant theoretical interest 
[e.g. @Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015], but such 
formulations start from the empirical observation of auto-correlation 
in environmental time series rather than a process-based derivation.
As the expansion clearly shows, any bounded continuous-time process will
involve auto-correlation -- measure a continuous process at ever-closer intervals in
time and these measurements must converge. As the derivation in Appendix B makes clear,
it is rather the the *relative* time scales of the environmental process
and population level process that really matter (more precisely, the relative 
magnitudes of the auto-correlation time of the environmental dynamics and population dynamics.)
An environment with comparatively short auto-correlation scale will act
effectively as white noise upon the population dynamics.

The van Kampen expansion for extrinsic factors (appendix B) also highlights
the importance of how environmental noise couples to the population dynamics.
A mechanistic description of environmental noise acts directly on parameters
of the ecological model: for instance, in our Levins' model we may imagine
that the extinction rate of patches does not occur at some fixed rate $e$
but rather at some stochastically fluctuating rate of a given mean, variance,
and auto-correlation.  The system size expansion illustrates how we replace
a model where stochasticity is nested inside some parameter into the simpler
format of the canonical equation where this contribution can be merely added
on as Gaussian noise at the end of a deterministic skeleton.  (The expansion
derives that the environmental noise is proportional to the square of the derivative
of the deterministic skeleton with respect to the environmental variable,
see appendix B.) Thus for the Levins' model, we can capture the fluctuations
introduced by demographic and environmental noise at equilibrium as
a Gaussian distribution with variance given by the sum of of demographic
and environmental contributions:

@@

where $(1-\frac{\bar{e}}{c}) N$ is the equilibrium population size.
This reflects the assumption of the canonical equation, Eq \eqref{canonical}
that the standard deviation (e.g. square root of the variance) scales in direct proportion to the system size $N$,
while the demographic noise scales as the square root.  Consequently,
for a large system, the demographic contribution becomes proportionally smaller
while the environmental contribution remains the same.

@@

This review has explored three paradigms in how noise is viewed throughout the 
ecological literature, which I have dubbed respectively: noise the *nuisance*, 
noise the *creator*, and noise the *informer*.  Noise can be seen as a nuisance
almost by definition: in examining the origins of noise, we have seen how 
stochasticity is introduced not because ecological processes are random in some fundamental
sense, but rather, because those processes are influenced by a complex combination
of forces we do not model explicitly.  In this view, noise captures all that additional
variation that is separate from the process of interest, and a rich array of 
statistical methods allow us to separate the one from the other in observations and
experiments.  By examining the origins of noise, we have seen that despite the complex
ways in this noise can enter a model, that a Gaussian white-noise approximation [@vanKampen2007; @Black2012] is often
approriate given a limit of a large system size -- a fact often invoked implicitly but
rarely derived explicitly from the theorems of @Kurtz1978 and others.

Building on these foundations, we turned to noise the creator, illustrating how
even small magnitude Gaussian noise could itself create and drive interesting phenomena.
While a purely statistical paradigm might look to explain patterns such as oscillations
or sudden transistions in terms of deterministic processes, this section highlighted how
noise can create and sustain cycles [e.g. @Nisbet1976; @Bjornstad2004] and switches [@Keeling2001].  While our examples focused on the most
tractable systems, a wealth of literature has explored such phenomena in ever more complex
contexts.  These examples paint a very different picture of noise, one where "everything matters" [@Bjornstad2001],
where it can be difficult to know what drives a pattern and where ommitting any of the
complexity (age or spatial structure, autocorrelation, individual heterogeneity) can
qualitatively alter the behavior of a model.

Our third paradigm seeks a more optimistic middle ground of noise the informer.
Here we saw the examples from an empricially driven literature on early warning signals
[@Scheffer2009; @Dai2012] view noise as a source of countless miniature experiments
which can reveal the underlying dynamics of a system and how they may be slowly changing.
In this context, noise does not act to create phenomena of interest directly. The 
sudden transistions we seek to anticipate are still explained by the deterministic part
of the model -- bifurcations. But nor is noise a nuisance that merely cloaks this
deterministic skeleton from plain view: rather, it becomes a novel source of information
that would be inaccessible from a purely deterministic approach.  I believe more examples
of how noise can inform on underlying processes is possible, but will require greater
dialog between these world views.

>>>>

---
title: "From noise to knowledge: how randomness generates novel phenomena and reveals information"
author:
  - name: "Carl Boettiger"
    affiliation: a
    email: "cboettig@berkeley.edu"
address:
  - code: a
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
#csl: trendsjournals.csl
csl: ecology-letters.csl
bibliography: refs.bib
output: 
  rticles::elsevier_article:
    keep_tex: true
  word_document: default
layout:  3p # review = doublespace, 3p = singlespace, 5p = two-column
preamble: |
  \newcommand{\ud}{\mathrm{d}}
  \usepackage[nomarkers]{endfloat}
  \journal{Ecology Letters}
  \linenumbers
  \usepackage{setspace}
  \doublespacing
abstract: |
  # Article Type: Review & Synthesis
  # Running title: From noise to knowledge
  # Figures: 4
  # Tables: 0
  # Word count: 7430
  # Word count in abstract: 181
  # Number of References: 111
  # Archive statement:
  All data (simulated data and code) is made available on GitHub (<https://github.com/cboettig/noise-phenomena>), and if accepted, will be archived with provided DOI in an appropriate scientific data repositoriy.

@@

@@

<!--
keywords:
  - stochasticity
  - demographic noise
  - environmental noise
  - colored noise
  - quasi-cycles
  - tipping points
-->

@@

In arguing for this shift, it essential to recognize this is a call for a bigger tent,
not for the rejection of previous paradigms.  What I will characterize as 'noise the nuisance'
reflects a predominately statistical approach, in which noise, almost by definition,
represents all the processes we are not interested in that create additional variation
which might obscure the pattern of interest.  By contrast, an extensive literature has
long explored how noise itself can create patterns and explain processes from population
cycling to coexistence.  These broad categories should be seen as a spectrum and not be
mistaken for either a sharp
dichotomy nor a reference to a strictly empirical-theoretical divide.  Each paradigm
expands upon rather than rejects the previous notion of noise: the recognition that
noise can create novel phenomena does not mean that noise cannot also obscure the
signal of some process of interest.  Likewise, seeking to use noise as a novel source
of information about underlying processes will be informed by both previous paradigms,
as our discussion will illustrate.

Accompanying this discussion, I provide concise and
commented code for simulating each of the models we will discuss as
Appendix A, and more mathematical background and derivations in Appendix B.
Numerical simulations permit poking and prodding
investigation unencumbered by either experimental design or mathematical formalism.
A copy of these appendices
is maintained at <https://github.com/cboettig/noise-phenomena>.

@@

It is often suggested that stochasticity is inherent or fundamental,
e.g. demographic stochasticity arises because the birth of an individual is probabilistic
[@Melbourne2012]. It is easy to forget that this is merely an assertion about a
model, not an assertion about reality [@Bartlett1960].  Only in quantum mechanics do we find
inherently probabilistic properties: births and deaths of organisms all have
far more mechanistic explanations which we simply summarize in statistical or average terms:
an individual death occurs at rate $\lambda$.
This is in fact an extrinsic factor (variables not modeled explicitly),
which makes birth appear probabilistic; (just as we describe the toss of a coin
as probabilistic when in fact it is a mechanistic outcome of Newton's laws).
Demographic stochasticity arises only subsequently, when we change scales from
a description at the individual level to one at the population level.

@@

Here, we will examine how this canonical model can emerge from a
lower-level description and how it can deviate from that description. This approach will also
set the stage to see how our understanding of stochastic models can be extended to better
reflect natural processes that shape not only for the so-called deterministic skeleton, but the noise terms as well.
In so doing, I will highlight two tools fundamental to this analysis, one computational and one analytical:
the exact simulation method of @Gillespie1977's stochastic simulation algorithm (SSA), and the
van Kampen system size expansion [@vanKampen1976].

@@

The @Gillespie1977 provides an exact algorithm
for simulating demographic stochasticity at an individual level, Figure \ref{gillespie}
and Appendix A.
The algorithm is a simple and direct implementation of the master
equation, progressing in random step sizes determined
by the waiting time until the next event.
Free from both the approximations and mathematical complexity,
the Gillespie algorithm is an interesting example of where we rely on a
numerical implementation to check the accuracy of an analytic approximation,
even in the case of simple models such as we will discuss.  Though the 
algorithm is often maligned as numerically demanding, it can be run much
more effectively even on large models on today's computers than when
it was first developed in the 70s, and remains an underutilized approach
for writing simple and approximation-free^[that is, free from the approximation
made by SDE models as we see in the van Kampen example.  All models are, of course,
only approximations.] stochastic ecological models.

@@

We choose this model because it has the same functional form
as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
but is explicit about how this net growth rate is divided between
birth and death rate.  This leads to real differences in observed stochastic properties
of the model. Both the numerical approach (details in Appendix A) and the 
analytical approach (Appendix B) illustrate that it is straight-forward
to use this same approach in alternate formulations.

Figure \ref{gillespie} shows the results of two exact SSA simulations of the classic patch model of @Levins1969,
with identical parameters in both panels except for the total number of available sites,
$N$, illustrating the magnitude of these fluctuations indeed scales in proportion
to $\sqrt{N}$ as postulated by the canonical equation, Eq \eqref{canonical}.

@@

Note that this result
is *different* than the canonical equation, Eq \eqref{canonical}, or more generally
the diffusion approximation [or Kramer-Moyal expansion @Kramers1940; @Moyal1949; @Gardiner2009]
which is much more typical in the ecological literature [e.g. @Nisbet1982; @Lande2003; 
@Ovaskainen2010]. @Black2012 gives an excellent introduction 
to the van Kampen expansion as a connection between individual-based and population-level
model descriptions and applications, but fails to point out these differences,
mistakenly characterizing the van Kampen expansion as synonymous with the diffusion approximation.
The diffusion approximation does not lead to a Gaussian distribution
(equivalently, an SDE or PDE with linear noise term, see Appendix B), but rather
a *non-linear* SDE of the form (e.g. @Ovaskainen2010)

@@

The van Kampen expansion can also be useful in illustrating how extrinsic
factors such as environmental noise enter into a stochastic model.
Like the expansion of demographic noise, this derivation leads to 
environmental noise that is approximately Gaussian in distribution,
but unlike the canonical model, leads to explicitly auto-correlated noise 
(see derivation, Appendix B).  Auto-correlated or colored environmental
noise has been a topic of significant theoretical interest 
[e.g. @Roughgarden; @Ellner; @Lawton; @Ripa1999; @Marshall2015], but such 
formulations start from the empirical observation of auto-correlation 
in environmental time series rather than a process-based derivation.
Any bounded continuous-time process will
involve auto-correlation -- measure a continuous process at ever-closer intervals in
time and these measurements must converge. Appendix B shows that
it is the *relative* time scales of the environmental process
and population level process that really matter (more precisely, the relative 
magnitudes of the auto-correlation time of the environmental dynamics and population dynamics.)
An environment with comparatively short auto-correlation scale will act
effectively as white noise upon the population dynamics.

The van Kampen expansion for extrinsic factors (appendix B) also highlights
the importance of how environmental noise couples to the population dynamics.
The system size expansion illustrates how we can replace
a model where stochasticity is nested inside some parameter into the simpler
format of the canonical equation where this contribution can be merely added
on as Gaussian noise at the end of a deterministic skeleton.  (The expansion
derives that the environmental noise is proportional to the square of the derivative
of the deterministic skeleton with respect to the environmental variable,
see appendix B.) Thus for the Levins' model, we can capture the fluctuations
introduced by demographic and environmental noise at equilibrium as
a Gaussian distribution with variance given by the sum of of demographic
and environmental contributions:

@@

where $(1-\frac{\bar{e}}{c}) N$ is the equilibrium population size.
This justifies the assumption of the canonical equation, Eq \eqref{canonical}
that the standard deviation (e.g. square root of the variance) scales in direct proportion to the system size $N$,
while the demographic noise scales as the square root.  Consequently,
for a large system, the demographic contribution becomes proportionally smaller
while the environmental contribution remains the same.

@@

This review has explored three paradigms in how noise is viewed throughout the 
ecological literature, which I have dubbed respectively: noise the *nuisance*, 
noise the *creator*, and noise the *informer*.  Noise can be seen as a nuisance
almost by definition: in examining the origins of noise, we have seen how 
stochasticity is introduced not because ecological processes are random in some fundamental
sense, but rather, because those processes are influenced by a complex combination
of forces we do not model explicitly.  In this view, noise captures all that additional
variation that is separate from the process of interest, and a rich array of 
statistical methods allow us to separate the one from the other in observations and
experiments.  By examining the origins of noise, we have seen that despite the complex
ways in this noise can enter a model, that a Gaussian white-noise approximation [@vanKampen2007; @Black2012] is often
appropriate given a limit of a large system size -- a fact often invoked implicitly but
rarely derived explicitly from the theorems of @Kurtz1978 and others.

Building on these foundations, we turned to noise the creator, illustrating how
even small magnitude Gaussian noise could itself create and drive interesting phenomena.
While a purely statistical paradigm might look to explain patterns such as oscillations
or sudden transitions in terms of deterministic processes, this section highlighted how
noise can create and sustain cycles [e.g. @Nisbet1976; @Bjornstad2004] and switches [@Keeling2001].  While our examples focused on the most
tractable systems, a wealth of literature has explored such phenomena in ever more complex
contexts.  These examples paint a very different picture of noise, one where "everything matters" [@Bjornstad2001],
where it can be difficult to know what drives a pattern and where omitting any of the
complexity (age or spatial structure, auto-correlation, individual heterogeneity) can
qualitatively alter the behavior of a model.

Our third paradigm seeks a more optimistic middle ground of noise the informer.
Here we saw the examples from an empirically driven literature on early warning signals
[@Scheffer2009; @Dai2012] view noise as a source of countless miniature experiments
which can reveal the underlying dynamics of a system and how they may be slowly changing.
In this context, noise does not act to create phenomena of interest directly. The 
sudden transitions we seek to anticipate are still explained by the deterministic part
of the model -- bifurcations. But nor is noise a nuisance that merely cloaks this
deterministic skeleton from plain view: rather, it becomes a novel source of information
that would be inaccessible from a purely deterministic approach.  I believe more examples
of how noise can inform on underlying processes is possible, but will require greater
dialog between these world views.
